Unbiased pangenome graphs

### 2022-09-08 manuscript/main.tex
add simplified asymptotic notation
stuff

The inductive proof of theorem \ref{thm:seqwish} demonstrates how to build a variation graph from sequences and their pairwise alignments.
However, a na\"{i}ve algorithm based on this model would require a very large amount of space.
Although our identifier space $|S|$ must include all of $Q$, in practice, we only store $S$, as $\bar{S}$ can be trivially computed.
Assume an all-to-all alignment of $N$ sequences in $A$ as input, and that all sequences are approximately identical, so that the induced variation graph has $|S|/N$ nodes.
The induction must maintain reference to all characters in all input sequences $O(|S|)$, all character-to-character matches $O(|A|) \approx O(|S| N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(|S|)$, the nodes of the graph $O(|\mathcal{V}|) \approx O(|S|/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(|S|)$, and the set of paths $O(|\mathcal{P}|) \approx O(|S|)$.
We also maintain the bitvector $\mathcal{X}$ to mark seen characters of $Q$ during graph induction, which requires $O(|S|)$ bits, equivalent to $O(|S| / \log_2 |S|)$ integer identifiers.
In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(|S|(N^2 + 1/N + 1/log_2{Q} + 4))$ space.

@@

To compact $\mathcal{G}$, we traverse $Q$, finding each entry in $\mathcal{Z}$ in turn, recording its start and end in $\mathcal{V}$, which can be understood as a character vector or string containing all the sequence in the nodes of $\mathcal{G}$.
%in a ``graph vector'' $\mathcal{G} = v_1,v_2 \ldots v_{|\mathcal{V}|}$ containing the ordered characters of the graph.
%Any position in $\mathcal{G}$ so marked must be the start or end of a node.
We subsequently use these markings to subdivide $\mathcal{V}$ into a compacted version $\mathcal{V}'$ where compacted node boundaries are marked in a auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{V}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
$\mathcal{B}$ allows us to compute compacted node ids using efficient rank operations \citep{Gog_2014}.
%such that $\forall_{i \in 1 \ldots |\mathcal{V}'|} |v_i| \ge 1$.
%In Algorithm \ref{alg:seqwish} we provide a detailed description of the graph induction algorithm.

>>>>

The inductive proof of theorem \ref{thm:seqwish} demonstrates how to build a variation graph from sequences and their pairwise alignments.
However, a na\"{i}ve algorithm based on this model would require a very large amount of space.
Although our identifier space $|S|$ must include all of $Q$, in practice, we only store $S$, as $\bar{S}$ can be trivially computed.
Assume an all-to-all alignment of $N$ sequences in $A$ as input, and that all sequences are approximately identical, so that the induced variation graph has $|S|/N$ nodes.
The induction must maintain reference to all characters in all input sequences $O(|S|)$, all character-to-character matches $O(|A|) \approx O(|S| N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(|S|)$, the nodes of the graph $O(|\mathcal{V}|) \approx O(|S|/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(|S|)$, and the set of paths $O(|\mathcal{P}|) \approx O(|S|)$.
We also maintain the bitvector $\mathcal{X}$ to mark seen characters of $Q$ during graph induction, which requires $O(|S|)$ bits, equivalent to $O(|S| / \log_2 |S|)$ integer identifiers.
In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(|S|(N^2 + 1/N + 1/log_2{Q} + 4))$ space, or simply $O(|S| N^2)$.

@@

To compact $\mathcal{G}$, we traverse $Q$, finding each entry in $\mathcal{Z}$ in turn, recording its start and end in $\mathcal{V}$, which can be understood as a character vector or string containing all the sequence in the nodes of $\mathcal{G}$.
%in a ``graph vector'' $\mathcal{G} = v_1,v_2 \ldots v_{|\mathcal{V}|}$ containing the ordered characters of the graph.
%Any position in $\mathcal{G}$ so marked must be the start or end of a node.
We subsequently use these markings to subdivide $\mathcal{V}$ into a compacted version $\mathcal{V}'$ where compacted node boundaries are marked in a auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{V}|$ such that the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
$\mathcal{B}$ allows us to compute compacted node ids using efficient rank operations \citep{Gog_2014}.
%such that $\forall_{i \in 1 \ldots |\mathcal{V}'|} |v_i| \ge 1$.
%In Algorithm \ref{alg:seqwish} we provide a detailed description of the graph induction algorithm.
